A function--sheaf dictionary for algebraic tori over local fields

### 2013-10-23 shanghai.tex
fiddling
Fixed a few botched instances of function--sheaf.

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[Function--sheaf dictionary for local tori]{A function--sheaf dictionary for algebraic tori over local fields}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
\classification{14L15 (primary), 11G25, 14F05, 14D24, 11S31 (secondary)}
\keywords{quasicharacters, quasicharacter sheaves, algebraic tori, Greenberg transforms, N\'eron models, local fields}

@@

In Part 1 of this paper we study the category $\QC(G)$ of quasicharacter sheaves on $G$.
Since $\QC(G)$ is a rigid monoidal category
the set of isomorphism classes $\QCiso{G}$ forms a group.
Motivated by the sheaf--function dictionary for commutative, connected
algebraic groups \cite{deligne:SGA4.5}*{sommes trig.}, 
we explore the trace of Frobenius morphism
\[
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\]
and show in Theorem~\ref{thm:snake} that it is an isomorphism, 
assuming that the geometric component group of $G$ is finitely generated.

In Part 2 we use Theorem~\ref{thm:snake} to make a sheaf--function dictionary
for algebraic tori over local fields.  Let $K$ be a non-archimedian local field
with finite residue field $\Fq$; we place no restriction on the characteristic of $K$.
Write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be a torus defined over $K$.  The N\'eron model $\TT$ of $T$
is a smooth group scheme over $\OK$ with generic fibre $T$ and a canonical isomorphism
$\TT(\OK) \cong T(K)$.  For each positive integer $n$, let $\GN{T}_n$ be the
Greenberg transform of $\TT \times_{\Spec{\OK}} \Spec{\OK/\pK^n}$; the commutative
group scheme $\GN{T}_n$ plays the r\^{o}le of $G$, above.  Theorem~\ref{thm:application}
allows us to translate characters $T(K)$ of depth less than $n$ into quasicharacter sheaves on $\GN{T}_n$:
\[
\QCiso{\GN{T}_n} 
\iso
\Hom_{< n}(T(K),\EEx).
\]  
We also describe a pro-group scheme $\GN{T}$ 
simultaneously geometrizing all quasicharacters ({\it i.e.}, finite-depth characters) of $T(K)$.

@@

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}*{sommes trig.} and many people have since referred to such objects
as character sheaves; see for example, \cite{kamgarpour:09a}. 
Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the sheaf--function dictionary to the case of certain commutative
group schemes which are not necessarily connected or finite type.  The main use
of the term character sheaf is of course due to Lusztig. It is applied to certain perverse
sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.

We close the introduction with a more detailed description of the structure of this paper.
In Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with certain properties.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the sheaf--function dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded and finite quasicharacter sheaves,
two subcategories of $\QC(G)$.  These alternate descriptions of quasicharacter sheaves play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} closes out Part \ref{part1} with the proof of the main theorem: trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

@@

\begin{thanks}

\end{thanks}

@@

When $G$ is connected, we may apply the classic sheaf--function dictionary.  In order to do so,
however, we must relate our notion of quasicharacter sheaf to the sheaf side of the standard dictionary.
Choose any $\Fq$-rational point of $G$ and let $\bg$ be the geometric point on $G$ lying above $g$.
Recall that the \emph{Weil group} of $G$, which we will denote by $\W(G,\bg)$, is a subgroup of the \'etale
fundamental group defined by the following diagram:
\[
 \begin{tikzcd}
 1 \rar & \ar[equal]{d} \pi_1(\bG, \bg) \rar & \W(G,\bg) \rar \dar[hook] & \Weil{} \rar \dar[hook] & 1 \\
 1 \rar &  \pi_1(\bG, \bg) \rar & \pi_1(G,\bg) \rar & \Gal(\bFq/\Fq) \rar & 1.
 \end{tikzcd}
\]
The $\Fq$-rational point $g$ under the geometric point $\bg$ determines a splitting
$\Weil{}\to \W(G,\bg)$ of $\W(G,\bg)\to \Weil{}$.

@@

  To show that $L_G(f,\psi) \iso (\qcs{L}_0,\mu_0)$ we simply apply the sheaf--function dictionary
  as follows \citelist{\cite{laumon:87a}*{1.1.3} \cite{deligne:SGA4.5}*{sommes trig.}}.
  First, recall that $L_G(f,\psi) \ceq (L(f,\psi), \mu(f,\psi))$.
  Let $k'/k$ be any degree-$n$ extension; 
  let $\Nm : G(k')\to G(k)$ be the norm map 
  and let $b : G_{k'} \to G$ be the base change map.
  By \cite{laumon:87a}*{1.1.3.3}, 
  $\trFrob{b^*L(f,\psi)} = \trFrob{\qcs{L}} \circ \Nm$;
  on the other hand, $\trFrob{\qcs{L}} \circ \Nm = \trFrob{b^*\qcs{L}}$,
   by Proposition~\ref{prop:basechange}.
  Using \cite{laumon:87a}*{Th\'eor\`eme 1.1.2}, it follows that  $L(f,\psi) \iso \qcs{L}_0$ in $\Loc(G)$.
  It is now clear that $(L(f,\psi),\mu(f,\psi)) \iso (\qcs{L}_0,\mu_0)$ in $\QCb(G)$.
  This completes the proof that $L_G : \QCf(G) \to \QC(G)$ is essentially surjective.
  Since we have already seen (in Section~\ref{sec:finite}) that $L_G$ is full and faithful, 
  it follows that $L_G$ is an equivalence.
\end{proof}

@@

\section{A function--function dictionary for tori over local fields} \label{sec:quasichar}

>>>>

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[A function--sheaf dictionary for local tori]{A function--sheaf dictionary for algebraic tori over local fields}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
\classification{14L15 (primary), 11G25, 14F05, 14D24, 11S31 (secondary)}
\keywords{quasicharacters, quasicharacter sheaves, algebraic tori, Greenberg transforms, N\'eron models, local fields, group schemes}

@@

In Part 1 of this paper we study the category $\QC(G)$ of quasicharacter sheaves on $G$.
Since $\QC(G)$ is a rigid monoidal category
the set of isomorphism classes $\QCiso{G}$ forms a group.
Motivated by the function--sheaf dictionary for commutative, connected
algebraic groups \cite{deligne:SGA4.5}*{sommes trig.}, 
we explore the trace of Frobenius morphism
\[
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\]
and show in Theorem~\ref{thm:snake} that it is an isomorphism, 
assuming that the geometric component group of $G$ is finitely generated.

In Part 2 we use Theorem~\ref{thm:snake} to make a function--sheaf dictionary
for algebraic tori over local fields.  Let $K$ be a non-archimedian local field
with finite residue field $\Fq$; we place no restriction on the characteristic of $K$.
Write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be a torus defined over $K$.  The N\'eron model $\TT$ of $T$
is a smooth group scheme over $\OK$ with generic fibre $T$ and a canonical isomorphism
$\TT(\OK) \cong T(K)$.  For each positive integer $n$, let $\GN{T}_n$ be the
Greenberg transform of $\TT \times_{\Spec{\OK}} \Spec{\OK/\pK^n}$; the commutative
group scheme $\GN{T}_n$ plays the r\^{o}le of $G$, above.  Theorem~\ref{thm:application}
allows us to translate characters $T(K)$ of depth less than $n$ into quasicharacter sheaves on $\GN{T}_n$:
\[
\QCiso{\GN{T}_n} 
\iso
\Hom_{< n}(T(K),\EEx).
\]  
We also describe a pro-group scheme $\GN{T}$ 
simultaneously geometrizing all quasicharacters ({\it i.e.}, finite-depth characters) of $T(K)$.

@@

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}*{sommes trig.} and many people have since referred to such objects
as character sheaves; see for example, \cite{kamgarpour:09a}. 
Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the function--sheaf dictionary to the case of certain commutative
group schemes which are not necessarily connected or finite type.  The main use
of the term character sheaf is of course due to Lusztig. It is applied to certain perverse
sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.

We close the introduction with a more detailed description of the structure of this paper.
In Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with certain properties.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded and finite quasicharacter sheaves,
two subcategories of $\QC(G)$.  These alternate descriptions of quasicharacter sheaves play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} closes out Part \ref{part1} with the proof of the main theorem: trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

@@

@@

When $G$ is connected, we may apply the classic function--sheaf dictionary.  In order to do so,
however, we must relate our notion of quasicharacter sheaf to the sheaf side of the standard dictionary.
Choose any $\Fq$-rational point of $G$ and let $\bg$ be the geometric point on $G$ lying above $g$.
Recall that the \emph{Weil group} of $G$, which we will denote by $\W(G,\bg)$, is a subgroup of the \'etale
fundamental group defined by the following diagram:
\[
 \begin{tikzcd}
 1 \rar & \ar[equal]{d} \pi_1(\bG, \bg) \rar & \W(G,\bg) \rar \dar[hook] & \Weil{} \rar \dar[hook] & 1 \\
 1 \rar &  \pi_1(\bG, \bg) \rar & \pi_1(G,\bg) \rar & \Gal(\bFq/\Fq) \rar & 1.
 \end{tikzcd}
\]
The $\Fq$-rational point $g$ under the geometric point $\bg$ determines a splitting
$\Weil{}\to \W(G,\bg)$ of $\W(G,\bg)\to \Weil{}$.

@@

  To show that $L_G(f,\psi) \iso (\qcs{L}_0,\mu_0)$ we simply apply the function--sheaf dictionary
  as follows \citelist{\cite{laumon:87a}*{1.1.3} \cite{deligne:SGA4.5}*{sommes trig.}}.
  First, recall that $L_G(f,\psi) \ceq (L(f,\psi), \mu(f,\psi))$.
  Let $k'/k$ be any degree-$n$ extension; 
  let $\Nm : G(k')\to G(k)$ be the norm map 
  and let $b : G_{k'} \to G$ be the base change map.
  By \cite{laumon:87a}*{1.1.3.3}, 
  $\trFrob{b^*L(f,\psi)} = \trFrob{\qcs{L}} \circ \Nm$;
  on the other hand, $\trFrob{\qcs{L}} \circ \Nm = \trFrob{b^*\qcs{L}}$,
   by Proposition~\ref{prop:basechange}.
  Using \cite{laumon:87a}*{Th\'eor\`eme 1.1.2}, it follows that  $L(f,\psi) \iso \qcs{L}_0$ in $\Loc(G)$.
  It is now clear that $(L(f,\psi),\mu(f,\psi)) \iso (\qcs{L}_0,\mu_0)$ in $\QCb(G)$.
  This completes the proof that $L_G : \QCf(G) \to \QC(G)$ is essentially surjective.
  Since we have already seen (in Section~\ref{sec:finite}) that $L_G$ is full and faithful, 
  it follows that $L_G$ is an equivalence.
\end{proof}

@@

\section{A function--sheaf dictionary for tori over local fields} \label{sec:quasichar}
